---
title: "Quake: An R Package for NOAA Data Visualization"
author: "Russell Politzky"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r echo=FALSE}
library(htmlTable)
```

## Introduction

`quake` provides functions to clean and visualize [NOAA](https://ngdc.noaa.gov/nndc/struts/form?t=101650&s=1&d=1) data.

`quake` supports two kinds of visualizations based on `ggplot2`; timelines
and maps.  Timelines show when earthquakes happened, while maps show where
they happened.

Timelines make use a number of aesthetics to show:

- the time at which each quake happened (x position on the timeline) and
- indicate the magnitude of each quake by the size of the marker used and
- indicate the number of associated deaths by the colour of the marker and
- indicate the country or region in which the quake ocurred by grouping
  these together on the y axis and,
- indicate the city/town where a quake occurred using labels pointing to the
  associated marker and,
- use transparency to indicate the degree of clustering in time, of earthquakes.

`quake` provides two custom `ggplot2` geometries to achieve this.
  
Maps show the positions of earthquakes and 

- indicate magnitude by marker size and
- show the location, magnitude and total, associated deaths by way of
  clickable pop-ups.

`quake` wraps functionality from the [leaflet](https://rstudio.github.io/leaflet/)
 package to achieve this.

The examples below show typical usage of all the functions exported 
by `quake`. Additional detail on each function is avavilable in the 
API documentation.

## Cleaning NOAA Data

The code below reads a sample file using `data.table's` `fread` function
and cleans it using `eq_clean_data`.  The sample shows the first three
rows and columns 1-5, and 48-49 of the cleaned output.

```{r clean_data}
library(data.table)
library(magrittr)
library(quake)

data_file_name <- system.file("extdata", "earthquakes.tsv", package = "quake")
clean_data     <- fread(data_file_name) %>% eq_clean_data

```
Rows 1-3 and columns 1-5 and 48-49 of the cleaned data are shown below.

```{r bits_of_clean_data}
htmlTable(x = clean_data[1:3, c(1:5, 48:49)]) # show 1st three rows and cols 1-5 and 48-49
```

`eq_clean_data` uses function `eq_location_clean`, which is also exposed 
by `quake`, to clean NOAA location information. The example below shows 
how to use `eq_location_clean` to produce a vector of clean, location 
strings.

```{r clean_locations}
library(data.table)
library(magrittr)
library(quake)

data_file_name  <- system.file("extdata", "earthquakes.tsv", package = "quake")
clean_locations <- fread(data_file_name)[, LOCATION_NAME] %>% eq_location_clean
```
The first few cleaned locations are shown below.

```{r head_locations, echo = FALSE}
htmlTable(
  x = data.frame(
    "Cleaned Location" = head(clean_locations)
  ),
  align = "l"
) # show 1st three rows and cols 1-5 and 48-49
```

## Plotting an Earthquake Timeline

`quake` has two custom, `ggplot2` geometries to plot earthquake data.
`geom_timeline` plots timelines while `geom_timeline_label` add 
labels to a timeline.

The example code below generates sample data and plots 
a timeline showing the dates, magnitudes and countries of the 
quakes.

### Generate sample data

```{r generate_data}
library(magrittr)
library(ggplot2)
library(data.table)
library(quake)

n        <- 20 # no samples
# get a list of countries and cities.
cty_file <- system.file("extdata", "countries_cities.csv", package = "quake")
ctry_cty <- (fread(cty_file))[sample(1:.N, size = n, replace = TRUE), ]

# Sample data
dt <- data.table(
  date      = as.Date('2017-01-01') + seq(1, 365, 365/n),
  country   = factor(ctry_cty$Country),
  location  = factor(ctry_cty$City   ),
  intensity = runif(n)*10,
  deaths    = runif(n)*12
)
```
The sample data `head`'s shown below.

```{r head_dt, echo = FALSE}
htmlTable(x = head(dt))
```

### Plot the sample data

- date on the x axis, 
- country on the y axis, 
- quake magnitude shown using marker size, 
- the number of associated deaths indicated by colour

```{r plot_data_multiple_ys, fig.width = 7}
dt %>%
  ggplot() +
  geom_timeline(
    aes(
      x    = date, 
      y    = country, 
      size = intensity, 
      col  = deaths
    ), 
    alpha = 0.8
  ) +
  labs(x = "DATE")                                     +
  scale_size_continuous (name = "Richter scale value") +
  scale_color_continuous(name = "# deaths"           ) +
  theme_classic()                                      +
  theme_timeline_with_y_axis_text
```

Notice `theme_timeline_with_y_axis_text` theme add-on, which is 
one of two themes provided by the `quake` package, the other being
`theme_timeline`, which turns off all y-axis graphics including
the text.

Also notice that by adding `y = country`, a timeline, separated
by country, is produced.  Without this optional aesthetic, all 
quakes will be plotted on a single timeline.

### Plot sample data without y axis

Here's the same plot with all quakes shown on one timeline, and 
without any y-axis decorations.

```{r plot_data_no_y_axis, , fig.width = 7}
dt %>%
  ggplot() +
  geom_timeline(
    aes(
      x    = date, 
      size = intensity, 
      col  = deaths
    ), 
    alpha = 0.8
  ) +
  labs(x = "DATE")                                     +
  scale_size_continuous (name = "Richter scale value") +
  scale_color_continuous(name = "# deaths"           ) +
  theme_classic()                                      +
  theme_timeline # no y axis
```

Note that in this example, no `y` aesthetic was used and that as such,
`geom_timeline` plots a single timeline showing all events.

## Adding Labels to An Earthquake Timeline

The example above shows a timeline without labels.  That is, 
the timeline doesn't indicate the actual locations at which 
the quakes happened.  The `geom_timeline_label` solves
this problem.

The example below shows a timeline, plotting the same 
data, but with labels added using `geom_timeline_label`.

```{r plot_data_labels, fig.width = 7}
dt %>%
  ggplot() +
  geom_timeline(       # plot the timeline
    aes(
      x    = date, 
      y    = country, 
      size = intensity, 
      col  = deaths
    ), 
    alpha = 0.8
  ) + 
  geom_timeline_label( # add labels.
    aes(
      label = location , # label text
      x     = date     , # x location
      y     = country  , # level to add label at 
      size  = intensity  #
    ),
    n_max = 5 # only label 5 most intense quakes
  ) +
  labs(x = "DATE")                                     +
  scale_size_continuous (name = "Richter scale value") +
  scale_color_continuous(name = "# deaths"           ) +
  theme_classic()                                      +
  theme_timeline_with_y_axis_text
```

## Locating Quakes on a Map

`quake` also maps NOAA data making it easy to identify fault lines.


```{r map_quakes, fig.width = 7, message = FALSE, warning = FALSE}
library(readr)
library(dplyr)
library(lubridate)
library(quake)

data_file_name <- system.file("extdata", "earthquakes.tsv.gz", package = "quake")

read_delim(data_file_name, delim = "\t")           %>% # read tsv, NOAA data
  eq_clean_data()                                  %>% # clean data
  filter(COUNTRY == "MEXICO" & year(date) >= 2000) %>% # filter to show Mexico
  mutate(popup_text = eq_create_label(.))          %>% # add col with HTML popups
  eq_map(annot_col = "popup_text")                     # render map with popup
```

